|

### 2022-05-17 article/article.Rmd
update memo and appendix
edits to article

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
date: \today
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves public goods delivery. We argue that changes in bargaining costs depend on whether additional legislators share the executive's party affiliation: More opposition members reduce the executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed 174 former city councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: legislature size; legislative bargaining; local public goods provision; regression discontinuity design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. In modern democracies, most countries have assemblies at
all levels of government. These institutions regulate the quantity and quality
of legislation, check and veto other political powers, and influence
policymaking. Moreover, legislatures also determine taxation, government
spending and oversee public programs \citep{auriol2012optimal,
north1989constitutions, poulsen2018political, stasavage2011states,
Weingastetal1981}.

@@

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
First, in mayor-council municipalities, mayors are the main responsible for
public service decisions, and like presidents and governors, they can boost
local public investments by altering the government budget. Second,
partisanship plays a key role in city politics, even in countries in which
ideological attachment is relatively weak. Party attachment structures
political views \citep{samuels2014power}, and local-level elites usually belong
to different factions and parties. They fight for power in the municipality by
controlling various parties and forming distinct pre-electoral
coalitions.\footnote{Elite political competition in the local level has also
been linked to violence and conflict in Brazil and elsewhere
\citep{hoelscher2015politics, mignozzetti2018predatory}.} In many cities,
family ties may be substitutes for the ideological component, but influential
local leaders usually sort themselves into different political groups
\citep{ferraz2020political, nunes1977coronelismo}.

@@

Our findings have implications that extend well beyond Brazil. A general form
of our argument is that it is easy for a politician to negotiate with members
of her in-group \citep{abbink2019, g9030057}. Specifically, several countries
have separate executive and legislative powers, such as the federal level of
presidential systems, state governments, and local city councils. As these
institutions are influenced by political allegiances, the dynamics we
expose here may be generalizable. Depending on legislative support, changing
the legislature size could increase the chance of electing politicians that are
not aligned with the executive, hence decreasing welfare. Despite the
particular effects on political support and welfare, the logic described in
this paper would remain relevant for empirically accessing the effect of
enlarged legislatures on welfare. Our findings may also help scholars
understand why other countries in Latin America and Asia can generate effective
governance despite having strong executives and large coalitions
\citep{pereira2012surprising}.

>>>>

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
date: \today
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves public goods delivery. We argue that changes in bargaining costs depend on whether additional legislators share the executive's party affiliation: More opposition members reduce the equilibrium public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality rates. To explore possible mechanisms, we surveyed 174 former city councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: legislature size; legislative bargaining; local public goods provision; regression discontinuity design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. In modern democracies, most countries have assemblies at
all levels of government. These institutions regulate the quantity and the
quality of legislation, check and veto other political powers, and influence
policymaking. Moreover, legislatures also determine taxation, government
spending and oversee public programs \citep{auriol2012optimal,
north1989constitutions, poulsen2018political, stasavage2011states,
Weingastetal1981}.

@@

In this paper, we propose a theory based on executive leadership and party
politics to explain how city council size impacts local public goods provision.
First, in mayor-council municipalities, mayors are the main responsible for
public service decisions and can boost local public investments by altering the
government budget. Second, partisanship plays a key role in city politics, even
in countries in which ideological attachment is relatively weak. Party
attachment structures political views \citep{samuels2014power}, and local-level
elites usually belong to different factions and parties. They fight for power
in the municipality by controlling various parties and forming distinct
pre-electoral coalitions.\footnote{Elite political competition in the local
level has also been linked to violence and conflict in Brazil and elsewhere
\citep{hoelscher2015politics, mignozzetti2018predatory}.} In many cities,
family ties may be substitutes for the ideological component, but influential
local leaders usually sort themselves into different political groups
\citep{ferraz2020political, nunes1977coronelismo}.

@@

Our findings have implications that extend well beyond Brazil. A general form
of our argument is that it is easy for a politician to negotiate with members
of her in-group \citep{abbink2019, alt2018giving}. Specifically, several countries
have separate executive and legislative powers, such as the federal level of
presidential systems, state governments, and local city councils. As these
institutions are influenced by political allegiances, the dynamics we
expose here may be generalizable. Depending on legislative support, changing
the legislature size could increase the chance of electing politicians that are
not aligned with the executive, hence decreasing welfare. Despite the
particular effects on political support and welfare, the logic described in
this paper would remain relevant for empirically accessing the effect of
enlarged legislatures on welfare. Our findings may also help scholars
understand why other countries in Latin America and Asia can generate effective
governance despite having strong executives and large coalitions
\citep{pereira2012surprising}.
